Bee visitation, pollination service, and plant yield in commodity and hybrid seed canola

### 2023-01-19 Manuscript/main.tex
Worked on revisions

% Since the processes linking insect visitation to seed production has several intermediate steps, and is contingent on resource factors, we need a more complete model of crop yield that links plant-level responses to biological processes that occur at the plot and field level.
% This work adds to the body of literature examining the resource context of pollination services \citep{haig1988} within cropping systems \citep{marini2015, tamburini2016, fijen2018, tamburini2019}, and identifies directions of future research in seed production systems.

% Our five main hypotheses are outlined as follows:
% \begin{enumerate} 
%     \item Foraging honey bees and leafcutter bees are sensitive to distance due to cost of flight \citep{schmid1985}, and deplete floral resources with visitation. Therefore, insect visitation should decline with distance from their hive or shelter to achieve equivalent rewards \citep{dukas1998, cresswell2000}. 
%     \item Visiting bees can deposit pollen on stigmas, so areas with high rates visitation should have more pollen deposition on stigmas than areas with low rates of visitation \citep{mesquida1988c, hoyle2007}, depending on foraging behaviour of the bee \citep{free1983}.
%     % \item Plant size is limited by resources, which can be reduced by competition with neighbours, so areas with fewer plants per m$^2$ should have larger plants \citep{angadi2003}. %Not revelvant
%     \item Seed production for a given plant is limited by the resources it is able to sequester. Therefore, large plants should have more resources at their disposal, and should produce more fruit per plant and produce a higher number of larger seeds per fruit \citep{galen1985, lawrence1993, marini2015}.
%     \item Seed production in resource-rich plants is also limited by pollination. Plants in areas with high pollen deposition should have a higher proportion of successful flowers \citep{sabbahi2005}, and a higher number of seeds per pod \citep{morandin2005}.
% \end{enumerate}

@@

\subsection*{Data collection}
From June through August of 2014 and 2015, we surveyed 29 commodity canola fields (14 in 2014, 15 in 2015) near Beaverlodge, Alberta and 31 fields (17 in 2014, 14 in 2015) near Lethbridge, Alberta (Figure \ref{fig:fieldMap}).
Commodity canola fields were selected based on the proximity of honey bee apiaries and site access. 
Colonies of Western honey bees (\emph{Apis mellifera} L., hereafter HB) were located at the corner or side of 28 of the 60 fields (stocking rates: 0.03 - 3.1 hives/ha) while 32 fields were unstocked. %range: 0.03 - 3.1 hives/ha, mean: 0.6, sd: 0.58 Apiaries were stationed at the edges of fields, often near to shrubs or tree-lines to act as a wind break.
Central-pivot irrigation systems were present at 14 of the 31 fields near Lethbridge.
Growers were also asked for canola variety information, but there was not enough replication to test for difference between varieties.
% Fields with central-pivot irrigation were mostly circular, but dryland (non-irrigated) fields were usually square or rectangular.

%Maps + climate graphs needed for publication
During 2015 and 2016, we also surveyed 35 hybrid seed canola fields (15 in 2015, 20 in 2016) near Lethbridge, Alberta, from June through August of each year, all of which had central-pivot irrigation, and were stocked with honey bee hives at a rate of 3.6 hives/ha (apiaries stationed in the corners of fields).
In seed fields, bays of hermaphroditic (hereafter ``male") and female plants are typically planted in 1- and 6-m wide linear strips (``bays"), respectively. 
Seed fields were stocked with shelters (1.3 - 2.6 shelters/ha, Figure \ref{fig:seedfieldPhoto}) containing alfalfa leafcutter bee cocoons (\textit{Megachile rotundata} F., hereafter LCB) at a density of 50-100,000 cocoons/ha. 
Both HBs and LCBs are managed pollinators that are not native to Alberta, and are uncommon outside of managed pollination systems and honey production yards, so we assumed that the hives and shelters were the sole sources of visiting bees in the observed fields.

% (1.3 or 2.6 shelters/ha).
% at a rate of approximately 2 shelters/ha
% Shelters used either ``standard" or ``double"  LCB stocking ($\approx$50,000 vs. 100,000 cocoons/ha), and shelters were placed at either ``standard" or ``double" densities (1.3 vs. 2.6 shelters/ha) in each field.
% Double stocking rates were only used in fields with double shelter density, creating three combinations of LCB stocking: ``standard", ``double shelter", and ``double shelter and double stocking".
%Woudick 5,6 = 64 shelters (single), Woudick 2 = 113 shelters (double + bees), Woudick 3 = 128 shelters Graham 1 = 135 shelters

% We recorded air temperature and wind speed before beginning observations (Table \ref{tab:weather}).
% \begin{table}[h]
%     \centering
%     \begin{tabular}{r|r|l|l|l|l}
%          Year & Area & Nectar volume & Nectar \% & Temperature & Wind Speed \\ \hline
%          2014 & Grande Prairie & 0.82 $\pm$ 0.035 & 56.1 $\pm$ 0.14 & 22.8 $\pm 0.9$ & 16.4 $\pm 2.0$  \\
%          2014 & Lethbridge & 0.85 $\pm$ 0.033 & 58.4 $\pm$ 0.15 & 24.7 $\pm 0.8$ & 4.6 $\pm 0.7$ \\
%          2015 & Grande Prairie & 0.63 $\pm$ 0.025 & 60.4 $\pm$ 0.22 & 21.7 $\pm 0.7$ & 26.7 $\pm 3.9$ \\
%          2015 & Lethbridge & 0.40 $\pm$ 0.021 & 58.4 $\pm$ 0.18 & 27.4 $\pm 0.7$ & 11.7 $\pm 1.5$ \\
%          2016 & Lethbridge & 0 & 0 & 23.5 $\pm 0.7$ & 6.2 $\pm 0.9$ \\
%     \end{tabular}
%     \caption{Nectar volume ($\mu$L), concentration (\%), air temperature (C) and wind speed (km/hr), averaged across all observations for each year and area. Nectar standing crop in seed canola was negligible in 2015 and 2016.}
%     \label{tab:weather}
% \end{table}

In both field types, we established sets of plots at varying distances from the nesting sources of pollinators. 
In commodity fields, 1 m$^2$ plots (271 total plots) were located at 5, 20, 100, and 500 m away from the field edge closest to the set of honey bee hives or potential sources of natural pollinating insects in unstocked fields (forests, shrublands, or grasslands). 
We observed very few wild bees in either field type (Table \ref{tab:propVisitors}), so these were excluded from the analysis.
In seed fields, we established plots at 5, 20, 100 (250 m in 2016), and 400 m into the field along a transect from the nearest set of honey bee hives, using pairs of plots at the edge of adjacent male and female bays (Figure \ref{fig:seedfieldPlots}).
To examine within-bay variation in visitation, we established a plot at the centre of the female bay at the 5m and 400m plot in each field. 
Distances to nearest shelters were measured using a Nikon\texttrademark Laser 800S Rangefinder.
Finally, we incorporated plot-level visitation data from Waytes (\citeyear{waytes2022}; same years and locations) to more accurately gauge the effect of distance from shelter on visitation rates (647 total plots).
All surveys occurred on fair-weather days (median temperature: 24.5$^\circ$C., range: 17--33$^\circ$C.) with no rain and minimal wind ($>$30 km/hr); wind speed and temperature were measured at each site with a Kestral\texttrademark 2500 weather meter.
% Mean air temperature was similar between years in Grande Prairie ($\approx$ 22$^\circ$C), but Lethbridge was warmer in 2015 (23 vs. 28$^\circ$C).
%Plots were marked with coloured stakes driven into the soil to locate them later in the season. %Full paper should have a table of climate variables.

@@

The underlying models of each pSEM were specified as generalized linear mixed-effects models (GLMMs).
GLMMs can model processes that occur at different levels; for example, we modeled seed size at the plant level, and pollen deposition at the flower level.
Random intercepts were used to model variance at the field and plot level, but plot-level variance terms had poor traces and low effective sample sizes, indicating low plot-level variance or insufficient sampling; therefore, we omitted plot-level random effects for all models (except for pollen count and pod set in seed fields, see Table \ref{tab:seedCoefs}).
Visitation models used $log$(observation length) as an ``offset" variable to account for differences in observation lengths.
We observed LCBs directly harassing HBs during the plot-level observations, attacking HBs both on the flowers and in the air (also seen by \citealp{batra1978} and \citealp{waytes2022}), but HBs were never the aggressor in these interactions, so we treated LCB visitation as a potential cause of HB visitation (Figure \ref{fig:allSEM}).
Pod set can suppress future vegetative growth and flower production \citep{stephenson1981}, but SEMs do not allow for causal loops \citep{grace2012}, so we included a path from pod set to flower production to approximate this process.
The direction of the path could be reversed, implying that large flower production reduces the proportion of pod set, but we considered only the first scenario, as \citet{sabbahi2006} showed that low pod set increases flower production.
Plant and pollen samples were not collected from male bays in seed fields, and there were no large differences in visitation rates for honey bees or LCBs (p = 0.32, 0.72, respectively) once other terms were included as covariates, so bay was excluded from further analysis.
See Equation sets S1 and S2 for GLMM specifications and Table \ref{tab:dataSummary} for a summary of the SEM variables.
% Random intercepts were used to model variance between fields, within-field (pollen, flower survival, seed count \& size), and within-plot (seed count \& size).
% Continuous responses were modeled using log-normal (plant density, plant size), square-root normal (flower density), and exponential-normal distributions (weight per seed), while counts were modeled as a negative-binomial (honey bee visits, pollen grains per stigma, flowers per plant, seeds per pod) or zero-inflated negative-binomial distribution (LCB visits) with a log-link function. (Move to appendix)
% Pod set (fraction of flowers that matured into pods) was modeled using a beta-binomial distribution with a logit-link function.

@@

% Our initial hypotheses for commodity fields:
% \begin{itemize}
% 	\item Honey bee visitation depends on distance from their hive
%     \item Pollen deposition depends on honey bee visitation
%     \item Seeds per pod depends on pollen deposition
%     \item Weight per seed depends on pollen deposition
%     \item Flower abortion rate depends on pollen deposition
%     \item Total yield depends on seeds per pod, weight per seed, and flower abortion rate
% \end{itemize}

% Our initial hypotheses for seed fields:
% \begin{itemize}
% 	\item Honey bee visitation depends on distance from edge of field (proxy for distance from hives), bay position, bay type, and distance to leafcutter shelter
%     \item leafcutter bee visitation depends on distance from their shelter, bay position, and bay type
%     \item Pollen deposition depends on honey bee and leafcutter visitation
%     \item Seeds per pod depends on pollen deposition
%     \item Weight per seed depends on pollen deposition and seeds per pod
%     \item Flower abortion rate depends on pollen deposition
%     \item Total yield depends on seeds per pod, weight per seed, and flower abortion rate
% \end{itemize}

@@

% HB visitation decreased with distance from the edge of the field in both commodity and seed canola, and increased with distance from LCB shelters; interestingly, LCB visitation also decreased with distance from the edge of the field in seed fields.
Honey bee flower visitation (per plot) decreased with distance away from the edge of the field in both commodity and seed canola fields.
In commodity fields, HB visitation decreased from 12 visits/hr per plot at the edge to 1.8 visits/hr at 400 m into the field (Figure \ref{fig:allVisits}a, apiary of 40 hives).
Lower stocking rates decreased HB visitation at the edge of the field to 7.4 visits/hr at 20 hives, while unstocked field edges had an average of 0.8 visits/hr.
Unsurprisingly (given the higher stocking rate of 160 hives per field), HB visitation was much higher in seed canola fields, and decreased from 96.4 visits/hr at edge to 85.7 visits/hr at the centre of the field (Figure \ref{fig:allVisits}a). 
HB visitation was also much lower near LCB shelters, dropping from 81 visits/hr at 20 m to 31 at 2 m (Figure \ref{fig:allVisits}b), likely due to competition or aggression from LCBs; the direct effect of LCB visitation on HB visitation was weak (p = 0.73, after accounting for LCB distance effect).

@@

LCB visitation increased pollen deposition in seed canola, but honey bees did not increase pollen deposition in commodity or seed canola (Figure \ref{fig:allPollen}a).
Pollen deposition on stigmas was high in commodity canola (mean: 293 grains/stigma, SD: 385, range: 0--3981), but honey bee visitation had no effect.
Pollen deposition declined weakly with distance from the edge of the field (p$=$0.09), but this only amounted to a 6\% average decrease (290 grains/stigma at the field edge vs. 240 at field centre, Figure \ref{fig:allPollen}b).
In seed canola, pollen deposition was much lower overall (mean: 22 grains/stigma, SD: 43, range: 0--578), and decreased from 26.3 grains/stigma at the edge to 11.5 at the centre of the field (Figure \ref{fig:allPollen}b), independent of bee visitation.
LCB visitation increased pollen deposition from 13 grains/stigma at 0 visits/hr to 25 pollen grains at 100 visits/hr (Figure \ref{fig:allPollen}a), but there was no effect of honey bee visitation (p$=$0.20), implying that most of the pollen deposition in seed fields is the result of LCB visitation.
Pollen deposition also decreased from 15 grains/stigma at the edge of the female bay to 9 at the centre of the bay independent of visitation (Figure \ref{fig:allPollen}c), suggesting that bees crossing over into the male bays may deposit less pollen as they move towards the centre.% \citep{pinnisch1990}.%, who likely visit flowers at the edge of the bay first. 
% Flies were also common visitors in commodity fields (Table \ref{tab:propVisitors}), but had no effect on pollen deposition (p$=$0.577), so only honey bees and LCBs were considered as potential pollinators (other flower visitors were rare).

@@

Plant size had by far the strongest positive effect on the number of pods produced (see large plant paths in Figure \ref{fig:allSEM}), and there was evidence of pod set suppressing flower production.
However, pollen deposition did not alter seed or pod production in commodity canola (Figure \ref{fig:allSeeds}a,b,c), nor was there a direct effect of honey bee visitation or distance from field edge, at least at the levels we observed.
In both commodity and seed canola, plant size had a strong positive effect on percent pod set and seeds per pod, but did not increase seed size in commodity canola (Figure \ref{fig:allSeeds}d,e,f).
Pod set in seed canola decreased with distance away from the field edge and LCB shelters, and was also lower at the centre of the female bay (independent of visitation and pollen deposition).
This distance effect was not seen in commodity canola, suggesting a more long-term pollen limitation effect in seed canola that was not captured by our snapshot of pollen and visitation.
Plant density also had a positive direct effect on seed size in both crops; this likely indicates field areas with better microclimate or soil resources rather than density effects \emph{per se}, as plants at the edges of the field or in ``field holes" are typically both smaller and less dense.
Finally, pod set also had a strong negative influence on flowers per plant in both crop types, indicating that high pod set suppresses future flower production.

@@

HB visitation in both field types declined with distance into the field, but overall visitation was much higher in seed fields than commodity fields, due to the higher HB stocking rate used in seed fields (3.6 vs. 0.6 hives/ha).
Since HBs travel to the crop from their hive outside the edge of the field \citep{robinson2022}, this decline was not surprising, but we did not expect a similar decrease in LCB visitation with distance into the field, as their shelters are located \textit{within} the field.
This decrease with distance into the field may have been caused by LCBs migrating from shelters at the centre of the field to the edge \citep{goerzen1995}, as female LCBs are central-place foragers and tend to not forage far from their nests \citep{peterson2005,pitts2011, brunet2019}.
However, LCBs are not as constrained to a single nest as HBs are, and can drift between shelters within a given field \citep{goerzen1995, pittsSinger2013}.
LCBs must also forage for leaf materials to create cocoons, but seem to prefer thinner leaves than canola, mainly plants in the families Fabaceae and Rosaceae \citep{sinu2018}.
Since seed canola fields have very few weeds within the crop, LCBs at the centre of the field may have few choices of leaf material, or alternative pollen and nectar sources.
Therefore, LCBs likely move from the shelters at which they were released to the shelters at the edge of the field, so as to gain access to better leaf material or more diverse pollen and nectar sources \citep{horne1995, horne1995b}.
% To test this, a mark-recapture study of female LCBs could be done to see if bees migrate to shelters at the edge of the field, or if visitation is higher due to other reasons (\textit{e.g.} feral populations migrating to edge of field, long-distance foraging from centre of field as in \citealp{peterson2005}).  %SH: FINE FOR THESIS, BUT CAN BE REMOVED FOR MANUSCRIPT
% This would explain why LCB visitation was higher at the edge of the field, and may also explain why this pattern was not seen in fields with half-stocked shelters.
% It would be interesting to see if there is a critical threshold of LCB density per-shelter that causes them to move to shelters at the edge of the field (similar to \citealp{pittsSinger2013}).
%HBs visited flowers at the edge of the field more frequently than those in the middle of the field, as predicted for central-place foragers with nesting aggregations outside the field (\citealp{olsson2014}, see Chapter 2).
%Similarly, the high concentration of canola nectar (\citealp{mohr1990}, 58.3\% brix from Chapter 2) may cause leafcutter bees to choose more diluted nectar under hot and dry conditions, similar to HBs \citep{ohguchi1983}. % canola pollen is valuable to bees \citep{mohr1988,huang2004},
% Potential future studies should investigate how leafcutter bees shift among shelters after emergence, how cocoon production in seed fields varies with distance, and should investigate inter-cropping leaf material plants (e.g. \emph{Polygonum}, \emph{Pisum}, or other large, thin-leaved plants) to encourage leafcutter bees to stay within the field.

@@

This study has revealed some of the detailed aspects of pollination in commodity and seed canola crops.
First, it shows how two important species of commercial pollinators decline with distance from their nesting sites, and how LCBs unexpectedly decline with distance from the edge of the field.
Second, it sheds light on the relative importance of HBs and LCBs as pollinators, showing that levels of HB stocking used for honey production do not strongly influence yield in commodity canola crops, and the importance of LCBs as primary pollinators of seed canola.
Finally, it contextualizes the value of pollination alongside plant resources, showing that plant resources and pollen limit production in seed canola, while only plant resources limit production in commodity fields.
Our unique SEM approach allowed us to formulate and test relationships within a causal framework, which we consider to be extremely valuable, and highly underused in agricultural studies.
Our model serves as an approximation of a dynamic process, but could be built upon by using dynamic linear models \citep{iwasa2000, nord2011, saez2018} to examine how plants dynamically respond to changes in pollination.
These would lend greater understanding to the process of hybrid seed production, and could be used to more accurately predict crop yields in novel scenarios.

@@

>>>>

@@

\subsection*{Data collection}
From June through August of 2014 and 2015, we surveyed 29 commodity canola fields (14 in 2014, 15 in 2015) near Beaverlodge, Alberta and 31 fields (17 in 2014, 14 in 2015) near Lethbridge, Alberta (Figure \ref{fig:fieldMap}).
Commodity canola fields were selected based on the proximity of honey bee apiaries and site access. 
Colonies of Western honey bees (\emph{Apis mellifera} L., hereafter HB) were located at the corner or side of 28 of the 60 fields (stocking rates: 0.03 - 3.1 hives/ha while 32 fields were unstocked. %range: 0.03 - 3.1 hives/ha, mean: 0.6, sd: 0.58 Apiaries were stationed at the edges of fields, often near to shrubs or tree-lines to act as a wind break.
Stocking rates were calculated using the area of the focal field, but since other canola fields (and wild flowering plants) in were in different stages of bloom around the focal field, the \emph{effective} stocking rates were likely lower.
Central-pivot irrigation systems were present at 14 of the 31 fields near Lethbridge.
Growers were also asked for canola variety information, but there was not enough replication to test for difference between varieties.
% Fields with central-pivot irrigation were mostly circular, but dryland (non-irrigated) fields were usually square or rectangular.

During 2015 and 2016, we also surveyed 35 hybrid seed canola fields (15 in 2015, 20 in 2016) near Lethbridge, Alberta, from June through August of each year, all of which had central-pivot irrigation, and were stocked with HB hives at a rate of 2.5 hives/ha (160 hives per quarter section, apiaries stationed in the corners or sides of field).
In seed fields, bays of hermaphroditic (hereafter ``male") and female plants are typically planted in 1- and 6-m wide linear strips (``bays"), respectively. 
Seed fields were stocked with shelters (1.3 - 2.6 shelters/ha, Figure \ref{fig:seedfieldPhoto}) containing alfalfa leafcutter bee cocoons (\textit{Megachile rotundata} F., hereafter LCB) at a density of 50-100,000 cocoons/ha. 
Both HBs and LCBs are managed pollinators that are not native to Alberta, and are uncommon outside of managed pollination systems and honey production yards, so we assumed that the hives and shelters were the sole sources of visiting bees in the observed fields.

In both field types, we established sets of plots at varying distances from the nesting sources of pollinators. 
In commodity fields, 1 m$^2$ plots (271 total plots) were located at 5, 20, 100, and 500 m away from the field edge closest to the set of HB hives or potential sources of natural pollinating insects in unstocked fields (forests, shrublands, or grasslands). 
We observed very few wild bees in either field type (Table \ref{tab:propVisitors}), so these were excluded from the analysis.
In seed fields, we established plots at 5, 20, 100 (250 m in 2016), and 400 m into the field along a transect from the nearest set of HB hives, using pairs of plots at the edge of adjacent male and female bays (Figure \ref{fig:seedfieldPlots}).
To examine within-bay variation in visitation, we established a plot at the centre of the female bay at the 5m and 400m plot in each field. 
Distances to nearest shelters were measured using a Nikon\texttrademark Laser 800S Rangefinder.
Finally, we incorporated plot-level visitation data from Waytes (\citeyear{waytes2022}; same years and locations) to more accurately gauge the effect of distance from shelter on visitation rates (647 total plots).
All surveys occurred on fair-weather days (median temperature: 24.5$^\circ$C., range: 17--33$^\circ$C.) with no rain and minimal wind ($>$30 km/hr); wind speed and temperature were measured at each site with a Kestral\texttrademark 2500 weather meter.
% Mean air temperature was similar between years in Grande Prairie ($\approx$ 22$^\circ$C), but Lethbridge was warmer in 2015 (23 vs. 28$^\circ$C).
% Plots were marked with coloured stakes driven into the soil to locate them later in the season. %Full paper should have a table of climate variables.

@@

The underlying models of each pSEM were specified as generalized linear mixed-effects models (GLMMs).
GLMMs can model processes that occur at different levels; for example, we modeled seed size at the plant level, and pollen deposition at the flower level.
Random intercepts were used to model variance at the field and plot level, but plot-level variance terms had poor traces and low effective sample sizes, indicating low plot-level variance or insufficient sampling; therefore, we omitted plot-level random effects for all models (except for pollen count and pod set in seed fields, see Table \ref{tab:seedCoefs}).
All visitation models used $log$(observation length) as an ``offset" variable to account for differences in observation lengths.
Commodity HB visitation models using stocking rate rather than number of hives fit the data equally well ($\Delta$AIC = 0.2) and did not change any other results; seed fields all had identical HB stocking rates, so this was not included as a covariate in seed HB visitation models.
We observed LCBs directly harassing HBs during the plot-level observations, attacking HBs both on the flowers and in the air (also seen by \citealp{batra1978} and \citealp{waytes2022}), but HBs were never the aggressor in these interactions, so we treated LCB visitation as a potential cause of HB visitation (Figure \ref{fig:allSEM}).
Pod set can suppress future vegetative growth and flower production \citep{stephenson1981}, but SEMs do not allow for causal loops \citep{grace2012}, so we included a path from pod set to flower production to approximate this process.
The direction of the path could be reversed, implying that large flower production reduces the proportion of pod set, but we considered only the first scenario, as \citet{sabbahi2006} showed that low pod set increases flower production.
Plant and pollen samples were not collected from male bays in seed fields, and there were no large differences in visitation rates for HBs or LCBs (p = 0.32, 0.72, respectively) once other terms were included as covariates, so bay was excluded from further analysis.
See Equation sets S1 and S2 for GLMM specifications and Table \ref{tab:dataSummary} for a summary of the SEM variables.
% Random intercepts were used to model variance between fields, within-field (pollen, flower survival, seed count \& size), and within-plot (seed count \& size).
% Continuous responses were modeled using log-normal (plant density, plant size), square-root normal (flower density), and exponential-normal distributions (weight per seed), while counts were modeled as a negative-binomial (HB visits, pollen grains per stigma, flowers per plant, seeds per pod) or zero-inflated negative-binomial distribution (LCB visits) with a log-link function. (Move to appendix)
% Pod set (fraction of flowers that matured into pods) was modeled using a beta-binomial distribution with a logit-link function.

@@

% Our initial hypotheses for commodity fields:
% \begin{itemize}
% 	\item HB visitation depends on distance from their hive
%     \item Pollen deposition depends on HB visitation
%     \item Seeds per pod depends on pollen deposition
%     \item Weight per seed depends on pollen deposition
%     \item Flower abortion rate depends on pollen deposition
%     \item Total yield depends on seeds per pod, weight per seed, and flower abortion rate
% \end{itemize}

% Our initial hypotheses for seed fields:
% \begin{itemize}
% 	\item HB visitation depends on distance from edge of field (proxy for distance from hives), bay position, bay type, and distance to leafcutter shelter
%     \item leafcutter bee visitation depends on distance from their shelter, bay position, and bay type
%     \item Pollen deposition depends on HB and leafcutter visitation
%     \item Seeds per pod depends on pollen deposition
%     \item Weight per seed depends on pollen deposition and seeds per pod
%     \item Flower abortion rate depends on pollen deposition
%     \item Total yield depends on seeds per pod, weight per seed, and flower abortion rate
% \end{itemize}

@@

% HB visitation decreased with distance from the edge of the field in both commodity and seed canola, and increased with distance from LCB shelters; interestingly, LCB visitation also decreased with distance from the edge of the field in seed fields.
HB flower visitation (per plot) decreased with distance away from the edge of the field in both commodity and seed canola fields.
In commodity fields, HB visitation decreased from 12 visits/hr per plot at the edge to 1.8 visits/hr at 400 m into the field (Figure \ref{fig:allVisits}a, apiary of 40 hives, i.e. 0.625 hives/ha in a 64 ha quarter-section).
Lower stocking rates decreased HB visitation at the edge of the field to 7.4 visits/hr at 20 hives, while unstocked field edges had an average of 0.8 visits/hr.
Unsurprisingly (given the higher stocking rate of 160 hives per field), HB visitation was much higher in seed canola fields, and decreased from 96.4 visits/hr at edge to 85.7 visits/hr at the centre of the field (Figure \ref{fig:allVisits}a). 
HB visitation was also much lower near LCB shelters, dropping from 81 visits/hr at 20 m to 31 at 2 m (Figure \ref{fig:allVisits}b), likely due to competition or aggression from LCBs; the direct effect of LCB visitation on HB visitation was weak (p = 0.73, after accounting for LCB distance effect).

@@

LCB visitation increased pollen deposition in seed canola, but HB visitation did not increase pollen deposition in commodity or seed canola (Figure \ref{fig:allPollen}a).
Pollen deposition on stigmas was high in commodity canola (mean: 293 grains/stigma, SD: 385, range: 0--3981), but HB visitation had no effect.
Pollen deposition declined weakly with distance from the edge of the field (p$=$0.09), but this only amounted to a 6\% average decrease (290 grains/stigma at the field edge vs. 240 at field centre, Figure \ref{fig:allPollen}b).
In seed canola, pollen deposition was much lower overall (mean: 22 grains/stigma, SD: 43, range: 0--578), and decreased from 26.3 grains/stigma at the edge to 11.5 at the centre of the field (Figure \ref{fig:allPollen}b), independent of bee visitation.
LCB visitation increased pollen deposition from 13 grains/stigma at 0 visits/hr to 25 pollen grains at 100 visits/hr (Figure \ref{fig:allPollen}a), but there was no effect of HB visitation (p$=$0.20), implying that most of the pollen deposition in seed fields is the result of LCB visitation.
Pollen deposition also decreased from 15 grains/stigma at the edge of the female bay to 9 at the centre of the bay independent of visitation (Figure \ref{fig:allPollen}c), suggesting that bees crossing over into the male bays may deposit less pollen as they move towards the centre.% \citep{pinnisch1990}.%, who likely visit flowers at the edge of the bay first. 
% Flies were also common visitors in commodity fields (Table \ref{tab:propVisitors}), but had no effect on pollen deposition (p$=$0.577), so only HBs and LCBs were considered as potential pollinators (other flower visitors were rare).

@@

Plant size had by far the strongest positive effect on the number of pods produced (see large plant paths in Figure \ref{fig:allSEM}), and there was evidence of pod set suppressing flower production.
However, pollen deposition did not alter seed or pod production in commodity canola (Figure \ref{fig:allSeeds}a,b,c), nor was there a direct effect of HB visitation or distance from field edge, at least at the levels we observed.
In both commodity and seed canola, plant size had a strong positive effect on percent pod set and seeds per pod, but did not increase seed size in commodity canola (Figure \ref{fig:allSeeds}d,e,f).
Pod set in seed canola decreased with distance away from the field edge and LCB shelters, and was also lower at the centre of the female bay (independent of visitation and pollen deposition).
This distance effect was not seen in commodity canola, suggesting a more long-term pollen limitation effect in seed canola that was not captured by our snapshot of pollen and visitation.
Plant density also had a positive direct effect on seed size in both crops; this likely indicates field areas with better microclimate or soil resources rather than density effects \emph{per se}, as plants at the edges of the field or in ``field holes" are typically both smaller and less dense.
Finally, pod set also had a strong negative influence on flowers per plant in both crop types, indicating that high pod set suppresses future flower production.

@@

HB visitation in both field types declined with distance into the field, but overall visitation was much higher in seed fields than commodity fields, due to the higher HB stocking rate used in seed fields (mean: 2.5 vs. 0.6 hives/ha).
Since HBs travel to the crop from their hive outside the edge of the field \citep{robinson2022}, this decline was not surprising, but we did not expect a similar decrease in LCB visitation with distance into the field, as their shelters are located \textit{within} the field.
This decrease with distance into the field may have been caused by LCBs migrating from shelters at the centre of the field to the edge \citep{goerzen1995}, as female LCBs are central-place foragers and tend to not forage far from their nests \citep{peterson2005,pitts2011, brunet2019}.
However, LCBs are not as constrained to a single nest as HBs are, and can drift between shelters within a given field \citep{goerzen1995, pittsSinger2013}.
LCBs must also forage for leaf materials to create cocoons, but seem to prefer thinner leaves than canola, mainly plants in the families Fabaceae and Rosaceae \citep{sinu2018}.
Since seed canola fields have very few weeds within the crop, LCBs at the centre of the field may have few choices of leaf material, or alternative pollen and nectar sources.
Therefore, LCBs likely move from the shelters at which they were released to the shelters at the edge of the field, so as to gain access to better leaf material or more diverse pollen and nectar sources \citep{horne1995, horne1995b}.
% To test this, a mark-recapture study of female LCBs could be done to see if bees migrate to shelters at the edge of the field, or if visitation is higher due to other reasons (\textit{e.g.} feral populations migrating to edge of field, long-distance foraging from centre of field as in \citealp{peterson2005}).  %SH: FINE FOR THESIS, BUT CAN BE REMOVED FOR MANUSCRIPT
% This would explain why LCB visitation was higher at the edge of the field, and may also explain why this pattern was not seen in fields with half-stocked shelters.
% It would be interesting to see if there is a critical threshold of LCB density per-shelter that causes them to move to shelters at the edge of the field (similar to \citealp{pittsSinger2013}).
%HBs visited flowers at the edge of the field more frequently than those in the middle of the field, as predicted for central-place foragers with nesting aggregations outside the field (\citealp{olsson2014}, see Chapter 2).
%Similarly, the high concentration of canola nectar (\citealp{mohr1990}, 58.3\% brix from Chapter 2) may cause leafcutter bees to choose more diluted nectar under hot and dry conditions, similar to HBs \citep{ohguchi1983}. % canola pollen is valuable to bees \citep{mohr1988,huang2004},
% Potential future studies should investigate how leafcutter bees shift among shelters after emergence, how cocoon production in seed fields varies with distance, and should investigate inter-cropping leaf material plants (e.g. \emph{Polygonum}, \emph{Pisum}, or other large, thin-leaved plants) to encourage leafcutter bees to stay within the field.

@@

This study has revealed some of the detailed aspects of pollination in commodity and seed canola crops.
First, it shows how two important species of commercial pollinators decline with distance from their nesting sites, and how LCBs unexpectedly decline with distance from the edge of the field.
Second, it sheds light on the relative importance of HBs and LCBs as pollinators, showing that levels of HB stocking used for honey production do not strongly influence yield in commodity canola crops, and the importance of LCBs as primary pollinators of seed canola.
Finally, it contextualizes the value of pollination alongside plant resources, showing that plant resources and pollen limit production in seed canola, while only plant resources limit production in commodity fields.
Our unique SEM approach serves as an approximation of a dynamic process, but could be extended with dynamic linear models \citep{iwasa2000, nord2011, saez2018} to examine how plants respond to changes in pollination over time.
These would lend greater understanding to the process of hybrid seed production, and could be used to more accurately predict crop yields in novel scenarios.

@@

\section{Acknowledgments}

This work was supported by the Alberta Crop Industry Development Fund (#20014C001R), the Alfalfa Seed Commission Fund, the Canadian Bee Research Fund, and the Alberta Beekeepers Commission.
We thank our tireless field assistants, Breannalee Beer, Zac Wagman, Chris Isaacs, Ian Johnson, Krista Fjordbotten, Jeff Kearns, Sam Knight, and Haley Shade. 
We also thank Megan Abarmayan, Shahid Majid, Mohamed Ibrahim, Ashley Londeau, and Joseph Desmarais, for their additional help with processing plants and flower stigmas, and Lynae Ovinge for her support in the lab.
Thanks to Ria van Giessen, Alfonso Parra, and Miki Miheguli at BASF for giving us access to their seed fields in 2015. 
Finally, many thanks to the farmers who gave us access to their land for this study, in particular: Dave Hofer, Ron Lamb, Dave McKee, Gord and James Graham, and many others.
